---
title: "Articulability"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sonority scoring}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(grams)
```

## Introduction

This vignette is unfinished.

<!-- 
  PROSE TO WRITE:
  - What is sonority scoring and why does it matter for anagram research?
  - The core problem: anagram candidates vary enormously in how word-like they
    sound, and selecting stimuli without controlling for this introduces noise.
  - How the sonority score is used in {grams}: feeds into compute_string_indices()
    and the Mahalanobis classifier.
-->

## The sonority hierarchy

Sonority scoring in {grams} is based on a six-level hierarchy, from least to most sonorous:

| Level | Class | ARPAbet symbols | Example |
|---|---|---|---|
| 0 | Plosive | T, P, K, B, D, G | [t] tip, [p] page |
| 1 | Fricative | F, TH, SH, CH, S, DH, ZH, V, JH, Z, HH | [f] fat, [s] sarah |
| 2 | Nasal | N, M, NG | [n] net, [m] man |
| 3 | Lateral | L | [l] live |
| 4 | Rhotic | R | [r] run |
| 5 | Vowel | IH, UH, EH, AW, AH, AE, AA, AY, AO, ER, EY, IY, OW, OY, UW | [i] it, [u] boot |

Glides (W, Y) are treated as vowels (level 5) in the hierarchy.

<!--
  PROSE TO WRITE:
  - Brief linguistic justification for the hierarchy ordering
  - How the hierarchy is used to compute the sonority score for a string
    (e.g. sonority sequencing principle, peaks and slopes)
  - What the raw score represents before penalty adjustment
-->

## Phonotactic rules

Pronounceability is also evaluated against a set of phonotactic rules for English
onsets and codas. Clusters are classified as either **illegal** (never occurring in
English) or **rare** (occurring in very few words).

### Onset rules

Onsets are evaluated after stripping any leading vowels. The following onset
exceptions are permitted regardless of other rules: `SPL`, `SPR`, `STR`, `SKR`, `SKW`.

| Status | Cluster | Notes |
|---|---|---|
| Illegal | NG | /ŋ/ is illegal word-initially |
| Illegal | SB, SD, SG | |
| Illegal | DL, TL | |
| Illegal | SR, ZR | |
| Illegal | ZB, ZD, ZG | |
| Illegal | NP, NGP, MT, NT, NK, ML | Nasal assimilation violations |
| Illegal | KN, GN | Archaic onsets not used in modern English |
| Rare | THR, SHR | |
| Rare | PW, BW | |

### Coda rules

Codas are evaluated after stripping any trailing vowels. The following coda
exceptions are permitted regardless of other rules: `NGTHS`, `MPST`.

| Status | Cluster | Notes |
|---|---|---|
| Illegal | SB, SD, SG | |
| Illegal | DL, TL | |
| Illegal | SR, ZR | |
| Illegal | PT, BD, GZ | |
| Illegal | HH | Illegal word-finally |
| Illegal | W, Y | Illegal word-finally |
| Rare | MP, NT | |
| Rare | LK, LD, RD, GD | |

<!--
  PROSE TO WRITE:
  - How illegal vs. rare status affects the sonority score (penalty weights)
  - Any linguistic sources or references for the rule set
-->

## Orthographic pre-screening

{grams} includes an orthographic pre-screening step that applies to strings
transcribed via grapheme-to-phoneme (g2p) conversion, where the transcription
may obscure illegal letter-level clusters.

<!--
  PROSE TO WRITE:
  - Explain the problem: g2p normalises orthography before transcription,
    so illegal letter clusters (e.g. "ngb") may not surface as illegal
    phoneme clusters in the transcription.
  - How the pre-screening works: consonant-only trigram frequencies
    (positional and non-positional) are used to penalise strings with
    low-frequency consonant clusters.
  - Why consonant-only trigrams: avoids false positives for pseudowords
    with unusual but legal vowel-containing trigrams (e.g. "cheab").
  - Known limitation: ngatu-type cases with illegal onsets followed
    immediately by a vowel cannot be caught because all their trigrams
    contain vowels.
-->

## Interpreting the sonority score

<!--
  PROSE TO WRITE:
  - What a high score means (more word-like, better sonority sequencing)
  - What a low score means (less pronounceable, more nonword-like)
  - Rough empirical ranges if known, or note that calibration is pending
  - How the score interacts with the Mahalanobis classifier
    (one of five features used to classify strings as word/pseudoword/nonword)
-->

## Known limitations

<!--
  PROSE TO WRITE:
  - Penalty weights for rare vs. illegal clusters are provisional and
    have not yet been empirically calibrated against pronounceability ratings.
  - The ngatu-type limitation (illegal onsets followed by vowels are not
    caught by the consonant trigram screen).
  - Any other edge cases or known failure modes worth documenting.
-->

## Full phoneme inventory

For reference, the complete set of ARPAbet phonemes and their sonority
classifications used in {grams}:

```{r sonority-table, echo = FALSE}
knitr::kable(
  grams:::sonority_table,
  col.names = c("Sonority level", "Class", "IPA", "Example", "ARPAbet"),
  align = "clcll"
)
```
